Topic modeling on EU and the USA research projects

### 2017-06-26 thesis_report/draft_thesis.tex
Fixed: mistakes in the report (all except Analysis)

Another funding agency that actively promotes the progress of science
is the National Science Foundation in the United States of America.
NSF supports research in all the non medical fields of science \&
engineering and aims to advance the national health and prosperity. It
funds approximately 24\% of the federally supported research
conducted by the USA's colleges and universities.\footnote{\url{https://www.nsf.gov/about/}}

@@

\section{Theoretic background}
\label{theoretic}
In this chapter we will present some basic theoretic background about
topic modeling. To deal with extensive text body \todo{probably
  plural}, machine learning researchers have developed topic modeling:
a suite of methods that analyze the words of the original text to
discover the themes that run through the data and detect hidden
semantic structures by analyzing a corpus. By marking up the documents
using these topics and by using the resulting structure, the model can
be used for organization, finding similar documents, or any other
similar tasks. A topic is here defined as a probability distribution
of words, meaning that certain words are more likely within a certain
topic. Since the topics emerge from the analysis of the original
texts, topic modeling algorithms do not require any prior annotations
or labeling of the documents.\cite{Blei11introductionto}

@@

where $\mathbb{E}$ is the expectation of the random variable $X$.
\footnote{Suppose random variable $X$ can take value $x_{1}$ with 
probability $p_{1}$, value $x_{2}$ with probability $p_{2}$, 
and so on, up to value $x_{k}$ with probability $p_{k}$. 
Then the expectation of this random variable $X$ is defined as
$\mathbb{E}[X] = x_{1}p_{1} + x_{2}p_{2} + ... + x_{k}p_{k}$.}

@@

Like any discrete distribution, a multinomial has to sum to one, where the sum
is over all possible data points. Here, a data point is a document containing
$n$ words.\cite{Huang_maximumlikelihood}

@@

First, we'll introduce some basic models for topic modeling. The
Latent Semantic Indexing, or LSI, was presented by Deerwester et al.
in 1990.\cite{Deerwester90indexingby} The model manages to deal with the
problem that multiple terms can refer to the same meaning, i.e.
synonymy. However, it is not as successful regarding polysemy. The
reason is that every term is represented as just one point in the
so-called latent semantic space. Furthermore, a word that can mean two
or more different things is represented as a weighted average of the
different meanings.

@@

Latent Semantic Indexing was one of the earliest methods for finding
relationships between documents and the words that occur in
them.\cite{Deerwester90indexingby} In LSI a term-document matrix is
created by analyzing the corpus, where the rows correspond to words
and columns to documents. Each element in this sparse matrix describes
the number of times a word occurs in a document but this term count
can also be weighted with for instance TF-IDF.\footnote{TF-IDF stands 
for term frequency - inverse document frequency, and the tf-idf weight is 
a weight often used in information retrieval and text mining. This weight 
is a statistical measure used to evaluate how important a word is to 
a document in a collection or corpus. The importance increases
proportionally to the number of times a word appears in the document (TF) but
is offset by the frequency of the word in the corpus (IDF).} Letting
each word represent a dimension in a very high dimensional space, a
document can be seen as a vector with components corresponding to its
weighted term counts. \cite{Salton:1988:TAA:54259.54260}

@@

Probabilistic Latent Semantic Indexing evolved from LSI and uses the same
concept of finding a lower rank approximation of the term-document occurrence
matrix. The difference is that instead of being based on linear algebra, PLSI
is based on a mixture decomposition using a latent class model.
~\cite{Hofmann:1999:PLS:312624.312649} PLSI associates an unobserved class 
variable $z$ with each document-word observation pair (\textbf{w},$
w)$. This $z$ can be seen as a topic, since it is a probability distribution
over words. As a generative model, PLSI can be defined in the following way 
for a corpus $D$:

\begin{enumerate}
\item Pick a document \textbf{w} with probability $p($\textbf{w}$)$
\item For each of the $N$ words in \textbf{w}:
\begin{description}
\item (a) Pick a topic $z$ with probability $p(z|$\textbf{w}$)$
\end{description}
\begin{description}
\item (b) Generate a word $w$ with probability $p(w|z)$
\end{description}
\end{enumerate}
\begin{center}
\begin{figure}[h]
\centering
\includegraphics[width=0.6\textwidth]{figs/PLSI_plate_notation.png}
\caption{Plate notation representing the PLSI model. \textbf{w} is the document 
index, $z$ is the topic drawn for the word $w$ from the topic distribution of 
the document}
\end{figure}
\end{center}
The result from PLSI is that every document is represented as mixing proportions
for the topics, given by $p(z|$\textbf{w}$)$. Even though PLSI is generative for 
the analyzed corpus, it is not generative for new documents, which means is that
there is no clear way of assigning probability to a document that is not part
of the training data.\cite{blei2003latent} Another problem is that the number 
of parameters in the model grows linearly with the number of documents.

\subsubsection{Latent Dirichlet Allocation (LDA)}
\label{LDAexplained}
LDA is a topic model that was first presented as a graphical model for topic
discovery by David Blei, Andrew Ng, and Michael I. Jordan in 2003. The basic
idea is that documents exhibit multiple topics. LDA is a generative statistical
model that allows sets of observations to be explained by unobserved groups
that explain why some parts of the data are similar. For example, if
observations are words collected into documents, it posits that each document
is a mixture of a small number of topics and that each word's occurrence is
attributable to one of the document's topics.\cite{Blei11introductionto}

A limitation with PLSI is that there is no generative model for these
proportions, making it difficult to handle unseen documents. The reduced
document description result in PLSI is a vector where each element describes the
mixing proportion for a topic. The Latent Dirichlet Allocation model tries to
solve this limitation by setting a Dirichlet prior on the topic
distribution.\cite{blei2003latent} In practice, this results in more reasonable
mixtures of topics in a document.

@@

The basic idea in LDA is that we define a topic to be a multinomial
distribution over a fixed vocabulary. Technically, the model assumes 
that the topics are generated first, before the
documents.\cite{Blei11introductionto} Now for each document in the
collection, we generate the words in a two-stage process:

@@

For each document $w$, in corpus $D$, the LDA generative process is as follows, 
with corresponding plate notation in Figure 2:

@@

A limitation with the standard LDA model is the fixed 
vocabulary. This means that new words cannot enter the model and that
you have to know a priori what words will model the corpus
effectively. A setting where this could be problematic is when
analyzing data where new words in the form of slang or names emerge
over time. Another limitation is the fact that one has to state the
number of topics the model should use. This is not a big issue, but
could lead to some of the resulting topics to be insignificant.
\begin{center}
\begin{figure}[h]
\centering
\includegraphics[width=0.6\textwidth]{figs/LDA_standard_model.png}
\caption{Plate notation of the standard LDA model}
\end{figure}
\end{center}

@@

The Expectation Maximization, or EM, algorithm is an approach to deriving an
algorithm for approximately obtaining maximum likelihood of variables in a
statistical model. Normally one obtains a maximum likelihood solution by taking
partial derivatives of the likelihood function with respect to all variables,
setting these derivatives to zero and solving the equations simultaneously.
\cite{Myung:2003}

@@

In LDA, our ultimate goal is to find the posterior distribution of
latent topic variables after observing training data. However,
computing this distribution is intractable, so we're forced to
approximate. One approximation approach is to use an optimization
method called Variational Bayes. In the original paper it's being used
a variational Bayes approximation of the posterior distribution;
\cite{blei2003latent} alternative inference techniques use 
MCMC with Gibbs sampling\cite{Griffiths02gibbssampling} and expectation 
propagation.\cite{Minka:2002:EGA:2073876.2073918} The goal of variational
inference is to approximate an intractable probability distribution
$p$, with a tractable one $q$, in a way that makes them as close as
possible given a similarity measure. In this thesis, variational
inference is used to approximate the posterior distribution of latent
variables. A new distribution over the hidden variables $q(z, \beta)$
(called the variational distribution) is introduced. This new
distribution has properties such that it can be efficiently
computed.\cite{Fox2011ATO}

@@

The specific updates are:
\begin{equation}
\phi_{d,word_{i},k} \propto e^{E_{q}(log \theta_{d,k})+E_{q}(log\beta_{k,word_{i}})}
\end{equation}
\begin{equation}
\gamma_{d,k} = \alpha + \sum_{word_{i}}\phi_{d,word_{i},k}n_{d,word_{i}}
\end{equation}
\begin{equation}
\lambda_{k,word_{i}}=\eta +\sum_{d}n_{d,word_{i}}\phi_{d,word_{i},k}
\end{equation}
Where $n_{d,word_{i}}$ is the number of occurrences of $word_{i}$ in
document $d$. The algorithm above has constant memory requirements and
empirically converges faster than collapsed Gibbs sampling. However,
it still requires a full pass through the entire corpus in each
iteration. It can be slow to apply to very large datasets.
\footnote{\url{https://wellecks.wordpress.com/2014/10/26/ldaoverflow-with-online-lda/}}

\paragraph{Online variational LDA}
\label{oLDAvb}
An online variational inference algorithm is proposed by Hoffman et all.\cite{onlineLDAvb} for fitting $\lambda$, the parameters to the variational
posterior over the topic distributions $\beta$. The "online LDA" is nearly as
simple as the variational algorithm, but converges much faster for large
datasets. Variational inference requires to pass through the whole dataset for
each iteration, and that can be a problem when working with big datasets and is
not an option to use cases where the data is arriving constantly. The Online
Variational Inference algorithm aims to solve both problems without compromising
the quality of the obtained topics and with an algorithm nearly as simple as
Variational Inference. For this algorithm we will only highlight the differences
respect the Variational Inference and the resulting algorithm. The key
difference is the way to fit the variational parameter $\lambda$ respect the
Variational Inference explained in the previous section.

@@

We can see that unlike variational LDA, in online LDA we only need to
hold a small chunk of the data at a time, and once we're done
analyzing it, we never need it again. As with batch, once we've
estimated $\lambda$, we can find the most probable words for each
topic by looking at the word probabilities in each row of
$\lambda$.\footnote{\url{https://wellecks.wordpress.com/2014/10/26/ldaoverflow-with-online-lda/}}

@@

The paper that mostly influenced our work is \emph{Finding Scientific Topics}, by
T. Griffiths and M. Steyvers.\cite{griffiths_steyvers04} They investigated
abstract data from articles published in the Proceedings of the National Academy
of Sciences (PNAS) from 1991 to 2001 and compared research topics obtained from
topic modeling with existing categories. In our implementation, we also use 
the abstracts of the projects as documents for the model.

@@

Inspired by Griffiths and Steyvers, L. Sun and Y. Yin wrote a paper,
\emph{Discovering themes and trends in transportation research using topic
modeling}, \cite{Sun201749} about topic modeling in transportation
research. Considering approximately 17000 abstracts of articles
published on 22 leading journals in transportation research as
documents, they infer 50 topics by using LDA. Compared to \emph{Finding
scientific topics}, Sun and Yin use more than one journal (22 popular
journals) and they investigate the journal-topic distribution and
journal similarity. In our analysis, we identify the EU and the USA as
2 regions (like 2 different journals), with different proportions of
documents-projects, and we visualize the topic similarity of the two
regions per framework programme.

@@

\section{Methodology}
\label{methodology}
In this section, we explain the procedure that we followed in order to
apply the LDA model in our dataset. It consists of information about 
EU projects funded for research and innovation under the FP4 - H2020 
and the awards/projects that got funded by NSF from 1994 - 2017 in USA. We
consider the projects' description/abstract as the documents in LDA.

@@

To obtain the collection of the documents, we consider as ``documents''
in LDA the project objectives, which is an accurate description for
the project. We also added the ``title' value of each project to the
objectives-documents, to increase the probability of finding accurate
topics. For the majority of the projects the ``start date' and ``end
date' value was null, so we kept the splitting of the dataset per
framework, in order to keep the correlation to the time. We followed
the same format in the dataset of USA awards.

\begin{center}
\begin{tabular}{l*{6}{c}r}
Framework & Years & No of docs & No of unique tokens & No of words in the corpus \\
\hline
FP4 & 1994-1998 & 14567 & 7293 & 364263 \\
FP5 & 1998-2002 & 17202 & 15756 & 1617232 \\
FP6 & 2002-2006 & 10091 & 15521 & 1437607 \\
FP7 & 2007-2013 & 25607 & 26499 & 3789837 \\
H2020 & 2014-2017 & 9055 & 15835 & 1342882 \\
\end{tabular}
\end{center}

As stopwords, we used the list that is provided by the Gensim Python
library\footnote{http://radimrehurek.com/gensim/} and we added following very
common words in the existing dataset: ``data'', ``based'', ``new'', ``project'',
``university'', ``student'', ``students'', ``research'', ``study'', ``program'',
``development'', ``study'', ``studies'', ``provide'', ``use''. We also removed 
the rare words that appear less than 5 times in the collection. 
Finally, we obtained a smaller corpus per framework (see the table above).

@@

The National Science Foundation (NSF) is a United States government
agency that supports fundamental research and education in all the
non-medical fields of science and engineering. Its medical counterpart
is the National Institutes of Health (ur data does not include 
material from NIH). The NSF funds approximately 24\% of all
federally supported basic research conducted by the United States'
colleges and universities. In some fields, such as mathematics,
computer science, economics, and the social sciences, the NSF is the
major source of federal backing.

@@

In the provided dataset, the information is splitted into XML files, separated
per year, from 1994 - 2017. In order to keep the correlation to time and be able
to compare the two datasets later, we splitted the projects based on the year
they were initialized (see Table below).

@@

Just like the EU dataset, we followed the same preprocessing steps to obtain the
final corpus. As stopwords, we used the list that is provided by the Gensim
Python library and we removed following very common words in the existing
dataset: ``data', ``work', ``based'', ``new'', ``project'', ``university'', ``student'',
``students'', ``research'', ``study'', ``program'', ``development'', ``study'', ``studies'', ``provide'', ``use'', ``understanding'', ``important'', ``support'', ``proposed''. We also removed the rare words that appear less than 5 
times in the collection. Finally, we obtained a smaller corpus per framework 
(see Table below).

\begin{center}
\begin{tabular}{l*{6}{c}r}
Framework& Years & No of docs & No of unique tokens & No of words in the corpus \\
\hline
FP4 & 1994-1997 & 36752 & 28038 & 4346604 \\
FP5 & 1998-2001 & 37674 & 30728 & 5166316 \\
FP6 & 2002-2006 & 54212 & 38476 & 8837298 \\
FP7 & 2007-2013 & 87341 & 48277 & 15673424 \\
H2020 & 2014-2017 & 35581 & 32268 & 7379219 \\
\end{tabular}
\end{center}

@@

In general, the most popular subject in European Union over the years
is Information systems. In our dataset we fount \todo{found} projects
related to almost all kinds of information systems, and especially
Quality management systems (1994-2002), Management information
systems, Network information systems and Information systems as a
service to all kinds of organizations, including universities and
companies. Since technology has an impact in every aspect nowadays, it
is normal to have such popularity over the years and it's becoming
popular in education, as well. The same period of time, Innovation in
Business emerged as a trending topic, with projects related to new
technologies and software systems in the industry and how they assist
the businesses and the market. In addition, until 2002 a very popular
subject related to Information systems, was Applications in mobile
networks, embedded and broadband systems, which nowadays has lost part
of its popularity in research.

@@

The third most popular subject is Molecular biology and Genetics.
These projects are related to the cellular molecules that carry out
the biological processes essential for the cell's functions and
maintenance. These affect many aspects of our lives, including
evolution, nutrition and the spread of diseases. \todo{This is not an
  essay, remove.} Especially from 1998 up to day, there is a focus in
Health care and Diagnostics, and for the last decade researchers are
studying the cancer detection and treatment, which is a common cause
of death in EU and the second most common cause of death in USA.
\todo{It seems now that you are discussing topics that are not popular
  throughout the period, but in the latest frameworks?}

Moreover, Environmental, Social and Economic Sustainability is a
subject very important to the research community and many researchers
work on this up to day. In short, sustainability takes into account
how we might live in harmony with the natural world around us,
protecting it from damage and destruction. \todo{This is not an essay,
  remove.} Many projects are related to one of the Three Pillars of
Sustainability\footnote{http
  ://www.thwink.org/sustain/glossary/ThreePillarsOfSustainability.htm}.
The European Union is considered by some to have the most extensive
environmental laws of any international organization, started by the
European Economic Community (EEC), which is the beginning of the EU's
environmental policy. Social and Economic policies are also popular in
the EU region over the last decade. Also, the last two decades there
is a focus on the Green energy production, like hydropower, solar and
nuclear energy.

@@

The National Science Foundation mostly supports fundamental research
and education in the non-medical science and engineering. Its medical
counterpart is the National Institutes of Health \todo{You have said
  so before.}.\footnote{\url{https://en.wikipedia.org/wik
    i/National_Science_Foundation}} With STEM being the top priority
in USA, \todo{How do you know that?} most of the projects founded by
NSF are related to engineering, computer science, mathematics and
information systems, in general. The subjects STEM \& Education,
Software engineering, Information systems and Mathematics, arise as
topics in all these years, from 1994-2017.

@@

From 1994-2006 and the last three years, International Research and
conferences came up as a trending topic. Re- searchers \todo{Re-
  searchers} have a great interest to collaborate with other
colleagues from all over the world and share their knowledge. A big
part of the projects funded by the EU that is related to international
research, is also related to Physics, in general, and High energy
physics in EU (see Appendix: Wordclouds of topics of the EU).
\todo{Are you talking about the EU, or the US here?}

NSF shows a great interest in international activities across all NSF
supported disciplines. The goal of these projects is to support high
quality research and education, which could not occur without
international collaboration. NSF seeks to catalyze a higher level of
international engagement in the US science and engineering community
by promoting international conferences and university projects and
partnerships \todo{You have said the same in the previous paragraph.}.
An interesting fact regarding international partnerships is that the
European Commission and the Government of the United States of America
concluded on 17 October 2016 an Implementing Arrangement that
simplifies the cooperation between Horizon 2020 projects and US
entities. European and American researchers will be able to work
together more closely on projects funded under Horizon 2020, the
European Union's research and innovation
programme.\footnote{\url{http://ec.europa.eu/research/iscp/index.cfm?pg=usa}}

@@

One of our goals is to identify trends in scientific research in EU
and USA, in order to give some insights to researchers and scientists
regarding potential trends. For this analysis, we run the model for
the whole dataset, instead of running the model per FP), for the EU
and the USA, respectively. The outcome was 10 topics for each region
from 1994 to 2017 for the EU and from 1994 to 2016 for the
USA.\footnote{in this implementation we removed the projects of the
  year 2017 from the USA dataset, because the amount of
  projects-documents was too small compared to the other years and it
  was affecting the result of the trend for each topic.}

>>>>

Another funding agency that actively promotes the progress of science
is the National Science Foundation in the United States of America.
NSF supports research in all the non medical fields of science \&
engineering and aims to advance the national health and prosperity. It
funds approximately 24\% of the federally supported research
conducted by the USA's colleges and universities\footnote{\url{https://www.nsf.gov/about/}}.

@@

\section{Theoretic background}
\label{theoretic}
In this chapter we will present some basic theoretic background about
topic modeling. To deal with extensive texts body, machine learning 
researchers have developed topic modeling:
a suite of methods that analyze the words of the original text to
discover the themes that run through the data and detect hidden
semantic structures by analyzing a corpus. By marking up the documents
using these topics and by using the resulting structure, the model can
be used for organization, finding similar documents, or any other
similar tasks. A topic is here defined as a probability distribution
of words, meaning that certain words are more likely within a certain
topic. Since the topics emerge from the analysis of the original
texts, topic modeling algorithms do not require any prior annotations
or labeling of the documents\cite{Blei11introductionto}.

@@

where $\mathbb{E}$ is the expectation of the random variable $X$
\footnote{Suppose random variable $X$ can take value $x_{1}$ with 
probability $p_{1}$, value $x_{2}$ with probability $p_{2}$, 
and so on, up to value $x_{k}$ with probability $p_{k}$. 
Then the expectation of this random variable $X$ is defined as
$\mathbb{E}[X] = x_{1}p_{1} + x_{2}p_{2} + ... + x_{k}p_{k}$.}.

@@

Like any discrete distribution, a multinomial has to sum to one, where the sum
is over all possible data points. Here, a data point is a document containing
$n$ words\cite{Huang_maximumlikelihood}.

@@

First, we'll introduce some basic models for topic modeling. The
Latent Semantic Indexing, or LSI, was presented by Deerwester et al.
in 1990\cite{Deerwester90indexingby}. The model manages to deal with the
problem that multiple terms can refer to the same meaning, i.e.
synonymy. However, it is not as successful regarding polysemy. The
reason is that every term is represented as just one point in the
so-called latent semantic space. Furthermore, a word that can mean two
or more different things is represented as a weighted average of the
different meanings.

@@

Latent Semantic Indexing was one of the earliest methods for finding
relationships between documents and the words that occur in
them\cite{Deerwester90indexingby}. In LSI a term-document matrix is
created by analyzing the corpus, where the rows correspond to words
and columns to documents. Each element in this sparse matrix describes
the number of times a word occurs in a document but this term count
can also be weighted with for instance TF-IDF\footnote{TF-IDF stands 
for term frequency - inverse document frequency, and the tf-idf weight is 
a weight often used in information retrieval and text mining. This weight 
is a statistical measure used to evaluate how important a word is to 
a document in a collection or corpus. The importance increases
proportionally to the number of times a word appears in the document (TF) but
is offset by the frequency of the word in the corpus (IDF).}. Letting
each word represent a dimension in a very high dimensional space, a
document can be seen as a vector with components corresponding to its
weighted term counts\cite{Salton:1988:TAA:54259.54260}.

@@

Probabilistic Latent Semantic Indexing evolved from LSI and uses the same
concept of finding a lower rank approximation of the term-document occurrence
matrix. The difference is that instead of being based on linear algebra, PLSI
is based on a mixture decomposition using a latent class model
~\cite{Hofmann:1999:PLS:312624.312649}. PLSI associates an unobserved class 
variable $z$ with each document-word observation pair (\textbf{w},$
w)$. This $z$ can be seen as a topic, since it is a probability distribution
over words. As a generative model, PLSI can be defined in the following way 
for a corpus $D$:

\begin{enumerate}
\item Pick a document \textbf{w} with probability $p($\textbf{w}$)$
\item For each of the $N$ words in \textbf{w}:
\begin{description}
\item (a) Pick a topic $z$ with probability $p(z|$\textbf{w}$)$
\end{description}
\begin{description}
\item (b) Generate a word $w$ with probability $p(w|z)$
\end{description}
\end{enumerate}
\begin{center}
\begin{figure}[h]
\centering
\includegraphics[width=0.6\textwidth]{figs/PLSI_plate_notation.png}
\caption{Plate notation representing the PLSI model. \textbf{w} is the document 
index, $z$ is the topic drawn for the word $w$ from the topic distribution of 
the document}
\end{figure}
\end{center}
The result from PLSI is that every document is represented as mixing proportions
for the topics, given by $p(z|$\textbf{w}$)$. Even though PLSI is generative for 
the analyzed corpus, it is not generative for new documents, which means is that
there is no clear way of assigning probability to a document that is not part
of the training data\cite{blei2003latent}. Another problem is that the number 
of parameters in the model grows linearly with the number of documents.

\subsubsection{Latent Dirichlet Allocation (LDA)}
\label{LDAexplained}
LDA is a topic model that was first presented as a graphical model for topic
discovery by David Blei, Andrew Ng, and Michael I. Jordan in 2003. The basic
idea is that documents exhibit multiple topics. LDA is a generative statistical
model that allows sets of observations to be explained by unobserved groups
that explain why some parts of the data are similar. For example, if
observations are words collected into documents, it posits that each document
is a mixture of a small number of topics and that each word's occurrence is
attributable to one of the document's topics\cite{Blei11introductionto}.

A limitation with PLSI is that there is no generative model for these
proportions, making it difficult to handle unseen documents. The reduced
document description result in PLSI is a vector where each element describes the
mixing proportion for a topic. The Latent Dirichlet Allocation model tries to
solve this limitation by setting a Dirichlet prior on the topic
distribution\cite{blei2003latent}. In practice, this results in more reasonable
mixtures of topics in a document.

@@

The basic idea in LDA is that we define a topic to be a multinomial
distribution over a fixed vocabulary. Technically, the model assumes 
that the topics are generated first, before the
documents\cite{Blei11introductionto}. Now for each document in the
collection, we generate the words in a two-stage process:

@@

For each document $w$, in corpus $D$, the LDA generative process is as follows, 
with corresponding plate notation in Figure \ref{figLDA}:

@@

A limitation with the standard LDA model is the fixed 
vocabulary. This means that new words cannot enter the model and that
you have to know a priori what words will model the corpus
effectively. A setting where this could be problematic is when
analyzing data where new words in the form of slang or names emerge
over time. Another limitation is the fact that one has to state the
number of topics the model should use. This is not a big issue, but
could lead to some of the resulting topics to be insignificant.
\begin{center}
\begin{figure}[h]
\centering
\includegraphics[width=0.6\textwidth]{figs/LDA_standard_model.png}
\caption{Plate notation of the standard LDA model}
\label{figLDA}
\end{figure}
\end{center}

@@

The Expectation Maximization, or EM, algorithm is an approach to deriving an
algorithm for approximately obtaining maximum likelihood of variables in a
statistical model. Normally one obtains a maximum likelihood solution by taking
partial derivatives of the likelihood function with respect to all variables,
setting these derivatives to zero and solving the equations simultaneously
\cite{Myung:2003}.

@@

In LDA, our ultimate goal is to find the posterior distribution of
latent topic variables after observing training data. However,
computing this distribution is intractable, so we're forced to
approximate. One approximation approach is to use an optimization
method called Variational Bayes. In the original paper it's being used
a variational Bayes approximation of the posterior distribution;
\cite{blei2003latent} alternative inference techniques use 
MCMC with Gibbs sampling\cite{Griffiths02gibbssampling} and expectation 
propagation\cite{Minka:2002:EGA:2073876.2073918}. The goal of variational
inference is to approximate an intractable probability distribution
$p$, with a tractable one $q$, in a way that makes them as close as
possible given a similarity measure. In this thesis, variational
inference is used to approximate the posterior distribution of latent
variables. A new distribution over the hidden variables $q(z, \beta)$
(called the variational distribution) is introduced. This new
distribution has properties such that it can be efficiently
computed\cite{Fox2011ATO}.

@@

The specific updates are:
\begin{equation}
\phi_{d,word_{i},k} \propto e^{E_{q}(log \theta_{d,k})+E_{q}(log\beta_{k,word_{i}})}
\end{equation}
\begin{equation}
\gamma_{d,k} = \alpha + \sum_{word_{i}}\phi_{d,word_{i},k}n_{d,word_{i}}
\end{equation}
\begin{equation}
\lambda_{k,word_{i}}=\eta +\sum_{d}n_{d,word_{i}}\phi_{d,word_{i},k}
\end{equation}
Where $n_{d,word_{i}}$ is the number of occurrences of $word_{i}$ in
document $d$. The algorithm above has constant memory requirements and
empirically converges faster than collapsed Gibbs sampling. However,
it still requires a full pass through the entire corpus in each
iteration. It can be slow to apply to very large datasets
\footnote{\url{https://wellecks.wordpress.com/2014/10/26/ldaoverflow-with-online-lda/}}.

\paragraph{Online variational LDA}
\label{oLDAvb}
An online variational inference algorithm is proposed by Hoffman et all.\cite{onlineLDAvb} for fitting $\lambda$, the parameters to the variational
posterior over the topic distributions $\beta$. The \emph{online LDA} is nearly as
simple as the variational algorithm, but converges much faster for large
datasets. Variational inference requires to pass through the whole dataset for
each iteration, and that can be a problem when working with big datasets and is
not an option to use cases where the data is arriving constantly. The Online
Variational Inference algorithm aims to solve both problems without compromising
the quality of the obtained topics and with an algorithm nearly as simple as
Variational Inference. For this algorithm we will only highlight the differences
respect the Variational Inference and the resulting algorithm. The key
difference is the way to fit the variational parameter $\lambda$ respect the
Variational Inference explained in the previous section.

@@

We can see that unlike variational LDA, in online LDA we only need to
hold a small chunk of the data at a time, and once we're done
analyzing it, we never need it again. As with batch, once we've
estimated $\lambda$, we can find the most probable words for each
topic by looking at the word probabilities in each row of
$\lambda$\footnote{\url{https://wellecks.wordpress.com/2014/10/26/ldaoverflow-with-online-lda/}}.

@@

The paper that mostly influenced our work is \emph{Finding Scientific Topics}, by
T. Griffiths and M. Steyvers\cite{griffiths_steyvers04}. They investigated
abstract data from articles published in the Proceedings of the National Academy
of Sciences (PNAS) from 1991 to 2001 and compared research topics obtained from
topic modeling with existing categories. In our implementation, we also use 
the abstracts of the projects as documents for the model.

@@

Inspired by Griffiths and Steyvers, L. Sun and Y. Yin wrote a paper,
\emph{Discovering themes and trends in transportation research using topic
modeling}\cite{Sun201749}, about topic modeling in transportation
research. Considering approximately 17000 abstracts of articles
published on 22 leading journals in transportation research as
documents, they infer 50 topics by using LDA. Compared to \emph{Finding
scientific topics}, Sun and Yin use more than one journal (22 popular
journals) and they investigate the journal-topic distribution and
journal similarity. In our analysis, we identify the EU and the USA as
2 regions (like 2 different journals), with different proportions of
documents-projects, and we visualize the topic similarity of the two
regions per framework programme.

@@

\section{Methodology}
\label{methodology}
In this section, we explain the procedure that we followed in order to
apply the LDA model in our dataset. It consists of information about 
EU projects funded for research and innovation under the FP4-H2020 
and the awards/projects that got funded by NSF from 1994-2017 in USA. We
consider the projects' description/abstract as the documents in LDA.

@@

To obtain the collection of the documents, we consider as ``documents''
in LDA the project objectives, which is an accurate description for
the project. We also added the ``title'' value of each project to the
objectives-documents, to increase the probability of finding accurate
topics. For the majority of the projects the ``start date'' and ``end
date'' value was null, so we kept the splitting of the dataset per
framework, in order to keep the correlation to the time. We followed
the same format in the dataset of USA awards.

\begin{center}
\begin{tabular}{l*{6}{c}r}
Framework & Years & No of docs & No of unique tokens & No of words in the corpus \\
\hline
FP4 & 1994-1998 & 14567 & 7293 & 364263 \\
FP5 & 1998-2002 & 17202 & 15756 & 1617232 \\
FP6 & 2002-2006 & 10091 & 15521 & 1437607 \\
FP7 & 2007-2013 & 25607 & 26499 & 3789837 \\
H2020 & 2014-2017 & 9055 & 15835 & 1342882 \\
\end{tabular}
\label{eutable}
\end{center}

As stopwords, we used the list that is provided by the Gensim Python
library\footnote{http://radimrehurek.com/gensim/} and we added following very
common words in the existing dataset: ``data'', ``based'', ``new'', ``project'',
``university'', ``student'', ``students'', ``research'', ``study'', ``program'',
``development'', ``study'', ``studies'', ``provide'', ``use''. We also removed 
the rare words that appear less than 5 times in the collection. 
Finally, we obtained a smaller corpus per framework (see table \ref{eutable}).

@@

The National Science Foundation (NSF) is a United States government
agency that supports fundamental research and education in all the
non-medical fields of science and engineering. Its medical counterpart
is the National Institutes of Health (our data does not include 
material from NIH). The NSF funds approximately 24\% of all
federally supported basic research conducted by the United States'
colleges and universities. In some fields, such as mathematics,
computer science, economics, and the social sciences, the NSF is the
major source of federal backing.

@@

In the provided dataset, the information is splitted into XML files, separated
per year, from 1994-2017. In order to keep the correlation to time and be able
to compare the two datasets later, we splitted the projects based on the year
they were initialized (see table \ref{usatable}).

@@

Just like the EU dataset, we followed the same preprocessing steps to obtain the
final corpus. As stopwords, we used the list that is provided by the Gensim
Python library and we removed following very common words in the existing
dataset: ``data'', ``work'', ``based'', ``new'', ``project'', ``university'', ``student'', ``students'', ``research'', ``study'', ``program'', ``development'', ``study'', ``studies'', ``provide'', ``use'', ``understanding'', ``important'', ``support'', ``proposed''. We also removed the rare words that appear less than 5 
times in the collection. Finally, we obtained a smaller corpus per framework 
(see Table \ref{usatable}).

\begin{center}
\begin{tabular}{l*{6}{c}r}
Framework& Years & No of docs & No of unique tokens & No of words in the corpus \\
\hline
FP4 & 1994-1997 & 36752 & 28038 & 4346604 \\
FP5 & 1998-2001 & 37674 & 30728 & 5166316 \\
FP6 & 2002-2006 & 54212 & 38476 & 8837298 \\
FP7 & 2007-2013 & 87341 & 48277 & 15673424 \\
H2020 & 2014-2017 & 35581 & 32268 & 7379219 \\
\end{tabular}
\label{usatable}
\end{center}

@@

In general, the most popular subject in European Union over the years
is Information systems. In our dataset we found projects
related to almost all kinds of information systems, and especially
Quality management systems (1994-2002), Management information
systems, Network information systems and Information systems as a
service to all kinds of organizations, including universities and
companies. Since technology has an impact in every aspect nowadays, it
is normal to have such popularity over the years and it's becoming
popular in education, as well. The same period of time, Innovation in
Business emerged as a trending topic, with projects related to new
technologies and software systems in the industry and how they assist
the businesses and the market. In addition, until 2002 a very popular
subject related to Information systems, was Applications in mobile
networks, embedded and broadband systems, which nowadays has lost part
of its popularity in research.

@@

The third most popular subject is Molecular biology and Genetics.
These projects are related to the cellular molecules that carry out
the biological processes essential for the cell's functions and
maintenance. Especially from 1998 up to day, there is a focus in
Health care and Diagnostics, and for the last decade researchers are
studying the cancer detection and treatment, which is a common cause
of death in EU and the second most common cause of death in USA.
\todo{It seems now that you are discussing topics that are not popular
  throughout the period, but in the latest frameworks?}

Moreover, Environmental, Social and Economic Sustainability is a
subject very important to the research community and many researchers
work on this up to day. 
The European Union is considered by some to have the most extensive
environmental laws of any international organization, started by the
European Economic Community (EEC), which is the beginning of the EU's
environmental policy. Social and Economic policies are also popular in
the EU region over the last decade. Also, the last two decades there
is a focus on the Green energy production, like hydropower, solar and
nuclear energy.

@@

With STEM being the top priority in USA, since this topic shows up in all the years
and according to the plot in figure \ref{usatrend} STEM is the topic with the highest proportion
of projects compared to the other topics (read section ), 
most of the projects founded by
NSF are related to engineering, computer science, mathematics and
information systems, in general. The subjects STEM \& Education,
Software engineering, Information systems and Mathematics, arise as
topics in all these years, from 1994-2017.

@@

From 1994-2006 and the last three years, International Research and
conferences came up as a trending topic. Researchers have a great 
interest to collaborate with other
colleagues from all over the world and share their knowledge. A big
part of the projects funded by the EU that is related to international
research, is also related to Physics, in general, and High energy
physics in EU (see Appendix: Wordclouds of topics of the EU).
\todo{Are you talking about the EU, or the US here?}

NSF shows a great interest in international activities across all NSF
supported disciplines. The goal of these projects is to support high
quality research and education, which could not occur without
international collaboration. NSF seeks to catalyze a higher level of
international engagement in the US science and engineering community
by promoting international conferences and university projects and
partnerships \todo{You have said the same in the previous paragraph.}.
An interesting fact regarding international partnerships is that the
European Commission and the Government of the United States of America
concluded on 17 October 2016 an Implementing Arrangement that
simplifies the cooperation between Horizon 2020 projects and US
entities. European and American researchers will be able to work
together more closely on projects funded under Horizon 2020, the
European Union's research and innovation
programme\footnote{\url{http://ec.europa.eu/research/iscp/index.cfm?pg=usa}}.

@@

One of our goals is to identify trends in scientific research in EU
and USA, in order to give some insights to researchers and scientists
regarding potential trends. For this analysis, we run the model for
the whole dataset, instead of running the model per FP), for the EU
and the USA, respectively. The outcome was 10 topics for each region
from 1994 to 2017 for the EU and from 1994 to 2016 for the
USA\footnote{in this implementation we removed the projects of the
  year 2017 from the USA dataset, because the amount of
  projects-documents was too small compared to the other years and it
  was affecting the result of the trend for each topic.}.

\subsection{Trends in the EU}
